---
title: "Project 1"
author: "Daniel Briseno and Emily Daskas"
date: "10/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#require("dplyr")
tips_df = read.csv("./Data/tips.csv")

attach(tips_df)
```


### Research Question 1: 

**Examine the association between the total bill and tip.** 

*How does this relationship change based on sex, smoker status, and the combination of the two (i.e. male smokers, female non-smokers, etc.)?*


```{r pressure, echo = FALSE}
outliers <- boxplot(tips_df$TIP/tips_df$TOTBILL, plot=FALSE)$out
tips_df <- tips_df[-which((tips_df$TIP/tips_df$TOTBILL) %in% outliers),]
plot(tips_df$TIP,tips_df$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main = "The Association between Tips and Total Bill")


```


"The Relationship between Tips and Total Bill" plot, seen above, shows a positive relationship between the tip amount and the total bill. This makes sense because people will generally tip somewhere around 15-20% of their total bill, so as their bill increases, their tip should too.

We can now look more in depth and see how this relationship changes based on smoking status and sex.


```{r, echo = FALSE}
smoker = tips_df[SMOKER==1,]
n_smoker =  tips_df[SMOKER==0,]
female = tips_df[SEX==1,]
male = tips_df[SEX==0,]

par(mfrow=c(1,2), oma=c(0,0,2,0))
smoker_plot = plot(smoker$TIP, smoker$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main= "Smokers")
n_smoker_plot = plot(n_smoker$TIP, n_smoker$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main= "Non-Smokers")
title("The Association between Tips and Total Bill Based on Smoking Status",outer = TRUE)
```
```{r, echo = FALSE}
boxplot((smoker$TIP/smoker$TOTBILL),(n_smoker$TIP/n_smoker$TOTBILL),
        names=c("smoker","nonsmoker"),main= "% Tip of Smokers vs Non-Smokers") 
```



In the above plots, we examine this positive relationship between tips and total bill based on smoking status. As seen in the scatter plot, when separating by these groups, there is a strong, positive correlation between tip amount and total bill for the non-smoking group. However, it seems as though that relationship is less clear in the smoking group. This could be because smoking is often associated with drinking. The relationship in the smoking group is less linear. We can also see in the box plot that the median tip percent of both the smokers and non-smokers is very similar, but the spread is much different. Non-smokers tip percent is much more consistent, while with smokers, you are likely to get a wider range of tips.


Next, we will take the same approach, but examine tipping habits based on sex.


```{r, echo = FALSE}

par(mfrow=c(1,2), oma=c(0,0,2,0))
female_plot = plot(female$TIP, female$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main= "Females")
male_plot = plot(male$TIP, male$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main= "Males")
title("The Association between Tips and Total Bill Based on Sex",outer = TRUE)
```

```{r, echo = FALSE}
boxplot((female$TIP/female$TOTBILL),(male$TIP/male$TOTBILL),
        names=c("female",'male'), main= "% Tip of Females vs Males") 
```

We can see in the plots above, that both males and females tip with similar habits. There is a positive, linear relationship between tips and total bill for both sexes. Interestingly, it seems as though females are morre likely going to tip slighly higher than men as the spreead of their inter quartile range is slighlty higher than the male category.

After examining males vs females and smokers vs non-smokers, we can go further in depth and look at a combination of both.

```{r, echo = FALSE}
m_smoker = tips_df[SEX==0 & SMOKER==1,]
f_smoker = tips_df[SEX==1 & SMOKER==1,]
m_nsmoker = tips_df[SEX==0 & SMOKER==0,]
f_nsmoker = tips_df[SEX==1 & SMOKER==0,]

par(mfrow=c(1,2), oma = c(4,4,2,0))
plot(m_smoker$TIP, m_smoker$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main = "Male")
plot(f_smoker$TIP, f_smoker$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main = "Female")
title("Association between Tips and Bill Among Smokers based on Sex",outer = TRUE)

par(mfrow=c(1,2), oma = c(4,4,2,0))
plot(m_nsmoker$TIP, m_nsmoker$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main = "Male")
plot(f_nsmoker$TIP, f_nsmoker$TOTBILL, xlab = "Tip Amount", ylab = "Total Bill", main = "Female")
title("Association between Tips and Bill of Non-Smokers based on Sex",outer = TRUE)
```


```{r, echo = FALSE}
boxplot((m_smoker$TIP/m_smoker$TOTBILL),(f_smoker$TIP/f_smoker$TOTBILL),
        names=c("male smoker","female smoker"), main= "% Tip of Male Smokers vs Female Smokers") 
boxplot((m_nsmoker$TIP/m_nsmoker$TOTBILL), (f_nsmoker$TIP/f_nsmoker$TOTBILL), 
        names=c("male nonsmoker","female nonsmoker"),main= "% Tip of Male Non-Smokers vs Female Non-Smokers") 

```


We find interesting differnces in the above graphs. First, by looking at smokers, we can see that females will more consistently tip higher than men. The median percent of men is 15% while it is about 18% for women. On the other hand, for non-smokers, we see that men have a higher tipping rate than women.  


### Research Question 2: 

**When are the best day and times to schedule a shift if you are looking to maximize tips?** 

```{r, echo = FALSE}

tips_df= read.csv("./Data/tips.csv")
# 
attach(tips_df)
tips_df = tips_df[, c("TIP", "DAY", "TIME", "TOTBILL")]
# # 
# # 

for(i in 1:ncol(tips_df)){
  tips_df[is.na(tips_df[,i]), i] <- mean(tips_df[,i], na.rm = TRUE)
}

outliers <- boxplot((tips_df$TIP), plot=FALSE)$out
tips_df <- tips_df[-which((tips_df$TIP) %in% outliers),]
# $TOTBILL) %in% outliers),]



tipsD = tips_df[TIME == 0,]
tipsN = tips_df[TIME == 1,]

thurD = tipsD[tipsD$DAY==3,]$TIP
thurN = tipsN[tipsN$DAY==3,]$TIP
friD = tipsD[tipsD$DAY==4,]$TIP
friN = tipsN[tipsN$DAY==4,]$TIP
satD = tipsD[tipsD$DAY==5,]$TIP
satN = tipsN[tipsN$DAY==5,]$TIP
sunD = tipsD[tipsD$DAY==6,]$TIP
sunN = tipsN[tipsN$DAY==6,]$TIP

data = c(thurD,thurN,friD,friN,satD,satN,sunD,sunN)
data_count = c(length(thurD),length(thurN),length(friD),length(friN),length(satD),length(satN),length(sunD),length(sunN))

colors = c("lightblue", "darkblue", "lightblue", "darkblue",  "lightblue", "darkblue", "lightblue", "darkblue")
spacing = c(1,2,4,5,7,8,10,11)
xLabels = c("Thursday, Day","Thursday, Night", "Friday, Day", "Friday, Night", "Saturday, Day", "Saturday, Night", "Sunday, Day", 'Sunday, Night') 

tip_shifts <- boxplot(thurD,thurN,friD,friN,satD,satN,sunD,sunN, 
        col = colors, 
        at= spacing, 
        names= xLabels,
        cex.axis = .8, 
        las = 2,
        outline=FALSE, main="Best Time to be Scheduled for a Shift",
      xlab="Shift", ylab="Tip Amount")
legend("topleft", legend = c("Day Shift","Night Shift"), 
    col = c("lightblue" ,"darkblue") , bty = "n", pch=20 , pt.cex = 2, cex = .6, horiz = FALSE, inset = c(0.03, 0.01))

```


The above boxplots tell us which shifts are optimal for if you desire to earn the most tips. This does not take into account the total bill amount, because if a server soley cares about tips, they will care less about how generous of a tip a party gives and more about the total tips they are able to generate during their shift, reguardless of how much a party's total bill is. 

It appears that the largest range, and total highest tips, comes in mostly on weekend nights. However, Friday and Saturday days have a higher median, but much lower maximum tips compared to their respective nights. This higher range of tips makes sense because people drink more on weekends. As we saw above, smokers, who are commonly associated with drinkers, have a larger spread of tip percentages and will often tip higher than non smokers.

The overall higher averges on weekends than weekday may be attributed to, aside from drinking, larger parties or more expensive means on weekends. Both of these contribute to larger total bills, which as we saw in our previous research question in our initial analysis, larger bill equate to larger tips.

Ultimately, as a server, if you are looking at the opportunity to make high tips while still guarunteeing a decent tip, Sunday night would be best, followed by Friday night. However, if you are looking for consistency in tipping, you will want to be scheduled for Thursday night or Saturday day as the spread is much tighter, but the median is much higher.




