---
title: "HW Lecture 7"
author: "Daniel Briseno Servin"
date: "11/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# R Shiny Meditation App #

For my feature, I was curious to see if users had definite preferences when it came to session types, and if there was any pattern to when users tended to not finish their sessions.

To this end I added 2 features. One was a new radio button which would toggle the scatterplot's shape mapping between session type and completed (yes or no). The purpose of this feature was to see if there was any relationship between time of day, date of session, or duration of session and whether or not the session was completed. The second feature I added was a bar chart below the scatterplot. This bar chart compares the number of sessions of a given type completed by a user. While the scatterplot's default shape mapping (session type) can show a user's preference between session types, this can be very hard to read, especially when there are many data points. The bar chart offers a more explicit comparison between session-types. I also added an option (in the form of a radio button) to only count completed sessions in the bar chart, which might show if there is a particular session type that users like to start, but often do not complete.

I found that there were not nearly enough incomplete sessions to come to any meaningful conclusions about their distribution, at least not with the implemented visual analysis. The bar chart was more informative. It showed that the most popular session types for each user tended to be meditation, music, and sleep -- often by a very wide margin. Among those three, meditation tended to be the most popular. Including or excluding incomplete sessions did not significantly change the bar charts.

**The app below can also be found [here](https://brise107.shinyapps.io/hwl7_rshiny/).**
```{r}
knitr::include_app("https://brise107.shinyapps.io/hwl7_rshiny/")
```




## App Code ##
```{r, echo=TRUE, eval=FALSE}
library(shiny)
library(dplyr)
library(reshape2)
library(ggplot2)
library(lubridate)

load("./Data/HW5-Data.RData")

#Data Pre-Processing####################################################

#Helper function, used to extract time of session from POSIX datetime
  #Input: POSIX datetime
  #Output: Minute count (numeric) of minutes past 00:00:00
as_min = function(t){
  hr = as.numeric(hour(t))*60
  return(as.numeric(minute(t))+ hr)
}

#Helper function, used to extract earliest and latest session POSIX timestamps for a given user ID
  #input User ID
  #Output (min,max) vector of earliest and latest logged_at_local timestamps
get_date_range = function(id){
  min_date = min((medtn %>% filter(ID==id))$logged_at_local)
  max_date = max((medtn %>% filter(ID==id))$logged_at_local)
  return(c(min_date, max_date))
}


#Split logged_at_local:POSIXct column into Day:chr (date) and Time:num columns
medtn_dt = medtn %>% mutate(day = date(logged_at_local),
                            time = as_min(logged_at_local) 
                            )
#Extract user ID's
id_nums  = medtn_dt$ID %>% unique()




#Define Page############################################################

ui <- fluidPage( titlePanel('Meditation Data'), window = 'Meditation Data',
                 sidebarLayout(
                   #Contains inputs
                    sidebarPanel(
                      #Drop Down Menu of Participant ID's
                      selectInput(
                        inputId = 'id',
                        label = 'Participant ID',
                        choices = id_nums
                      ),
                      br(),
                      #Date Range Select
                      dateRangeInput(
                        inputId = 'dates',
                        label = 'Date Range'
                      ),
                      br(),
                      #Shape-mapping Radio Button Select
                      radioButtons('shp', 
                                   'Select Shape Mapping', 
                                   choices = c('Session Type', 'Session Completed'),
                                   selected = 'Session Type' ),
                      br(),
                      radioButtons('complete_only',
                                   'Exclude incomplete sessions in Bar Plot?',
                                   choices = c('Yes', 'No'),
                                   selected = 'Yes')
                    ),
                    #Displays output
                    mainPanel(
                      #Scatter plot
                      plotOutput('dateVsStart_plt', brush = brushOpts(id="brsh")),
                      br(),
                      #text summary of brush-selected area from scatter plot
                      textOutput('selected_summary'),
                      br(), br(),
                      plotOutput('type_preference_bar')
                    )
                 )
  )

server <- function(input, output, session){
  
  #filter meditation dataframe by chosen participant ID
  filtered = reactive({ medtn_dt %>% filter(ID == input$id) })
  
  #update start and end dates for dateRangeInput (inputId='dates')
  observe({
    rng = get_date_range(input$id)
    updateDateRangeInput(session, 'dates',label = 'Date Range', start = rng[1], end = rng[2])
  })
  
  
  #define Date vs Start Time scatter plot output
  output$dateVsStart_plt <- renderPlot({
    plot_mappings = NULL
    
    #set scatterplot color and shape aesthetic maps
    if(input$shp == 'Session Type'){
        plot_mappings = geom_point(aes(shape = session_type, color = duration_in_sec ), size=3) 
    }
    else{
        plot_mappings = geom_point(aes(shape = completed, color = duration_in_sec ), size=3) 
    }
    
    #finish defining scatterplot
    ggplot(data = filtered(), aes(y=time,x=day)) + 
      plot_mappings +
      #set y-axis breakpoints and labels
      scale_y_continuous(breaks = seq(0,1440,by=240), labels = c('12am','4am','8am','12pm','4pm','8pm','12am'))+
      #Set y-axis and x-axis labels
      labs(y="Start Time", x = "Date") +
      #Change colors for duration_in_sec mapping
      scale_color_gradient(low = 'yellow', high = 'orangered')+
      xlim(input$dates[1],input$dates[2])
  })
  
  #define brush-selected text output
  output$selected_summary = renderText({
    num_selected = nrow(brushedPoints(filtered(),input$brsh))
    
    #default text output
    if(num_selected == 0){
      return("Please click and drag on the graph to select data")
    }
    #text output when user selects populated area on scatterplot
    else{
      
      selected = brushedPoints(filtered(),input$brsh)
      session_count  = nrow(selected)
      ave_session_length = mean(selected$duration_in_sec)/60
      ave_session_length = round(ave_session_length, digits = 2)
      
      format(c("There were ", session_count, " total observations selected. ",
        "The average duration for these observations is ", ave_session_length," min."))
    }
  })
  
  #Define barchart of session_types (height corresponds to number of sessions)
  output$type_preference_bar = renderPlot({
    session_type_data = filtered()
    #check if only completed sessions should be counted
    if(input$complete_only == 'Yes'){
      session_type_data <- session_type_data %>% filter(completed == "true")
    }
    ggplot(data=session_type_data, aes(session_type)) + geom_bar(stat='count') + 
      labs(y="Number of Sessions", x= "Session Type")
  })
}

shinyApp(ui=ui, server=server)

```

