---
title: "HW Lecture 3"
author: "Daniel Briseno Servin"
date: "9/21/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(knitr)
```

## Problem 1

*Merge the data frame from Task 5 with coords via three different techniques: inner, right, and left merge.  Compare the dimensions of each of the resulting data frame and explain any differences you see. *

```{r Q1}
#create dataframe from task 5
ccA = read.csv("./Data/countrycharsA.csv")
ccB = read.csv("./Data/countrycharsB.csv")
gdp = read.csv("./Data/gdp.csv")
ccAB = rbind(ccA,ccB)
ccAB_gdp = cbind(ccAB, gdp)

#load in coords dataframe
load("./Data/map.coords.RData")

#do inner, outer, left and right joins 
countryData_inner = merge(ccAB_gdp, coords, by = "country")
countryData_outer = merge(ccAB_gdp, coords, by = "country", all = T)
countryData_left = merge(ccAB_gdp, coords, by = "country", all.x=T)
countryData_right = merge(ccAB_gdp, coords, by = "country", all.y=T)

#compare dimensions

dim(countryData_inner)
dim(countryData_outer)
dim(countryData_left)
dim(countryData_right)
```
Here we can see that the outer join had the most rows, the inner had the least, and the left and right joins have a very close number of rows, with the left being greater by 20 rows.

This is what one would expect, since:

  - The outer join must include all entries from both tables, including the ones with no matching entry in the "countries" column. This means we can expect this one to have the most entries since it is the least exclusive
  - The inner join must include only the entries from each table which can be matched by the "country" attribute. We can expect this join to have the least entries since it is the most exclusive
  - The left join must include only the entries from `coords` which can be matched by "country" attribute to some entry in `ccAB_gdp`, and all entries of `ccAB_gdp`.
  - The right join must include only the entries from `ccAB_gdp` which can be matched by "country" attribute to some entry in `coords`, and all entries of `coords`.

The small difference between the left join and the right join tells us that there are more entries in `ccAB_gdp` which cannot by matched by country than there are in `coords`.

## Problem 2
*Suppose a researcher has a hypothesis that there is a relationship between gdp and distance from the equator, i .e. latitude. To explore this, the researcher would like to break   the gdp into quartiles and then look at the mean latitude (in absolute value) for each quartile. To do this, perform the following tasks:*

### A)
*Using  the  data  from  the  inner  merge,  use  the quantile() function  to  determine  the  quartiles of gdp*

```{r Q2a}
q = quantile(countryData_inner$gdp)
q
```

### B)
*Using  the  findInterval() function  and  your  result  from  above,  create  a  new factor variable  in  the  data  set  called gdp.   q  based  on  the  quartile  that  the  a  given  observation’s   gdp value lies in. Print a table of the levels   of this variable. Does your function seem to be working? How can you tell?*

```{r Q2b}
interval = findInterval(countryData_inner$gdp, q, all.inside = T)
gdp.q = factor(interval)
gdp.q_levels = data.frame(levels(gdp.q))
colnames(gdp.q_levels) = c("Quartile")
kable(gdp.q_levels)

```
The function does seem to be working, since we expect that with 5 quantiles the data will be partitioned into 4 quartiles, which we enforced by specifying `all.inside = T`. I believe that this worked, since we see 4 levels in the data which likely correspond to 4 quartiles.

### C)
*Reverse the order of the factor levels for gdp.q and reprint the table.*

```{r Q2c}
gdp.q = factor(gdp.q, levels = rev(levels(gdp.q)))
gdp.q_levels = data.frame(levels(gdp.q))
colnames(gdp.q_levels) = c("Quartile")
kable(gdp.q_levels)
```

### D)
*Find the  mean latitude (in absolute value) for  each  quartile.  Does  there  seem  to  be  a  difference?*

```{r Q2d}
countryData_q = cbind(countryData_inner,"gdpQuartile" = gdp.q)

q1_latitude = mean(abs(countryData_q[countryData_q$gdpQuartile == 1, "lat"]))
q2_latitude = mean(abs(countryData_q[countryData_q$gdpQuartile == 2, "lat"]))
q3_latitude = mean(abs(countryData_q[countryData_q$gdpQuartile == 3, "lat"]))
q4_latitude = mean(abs(countryData_q[countryData_q$gdpQuartile == 4, "lat"]))

mean_latitudes = c("Q1" = q1_latitude, "Q2" = q2_latitude,"Q3" = q3_latitude, "Q4" = q4_latitude)
mean_latitudes
```
The mean latitudes do not show a significant difference between the latitudes of countries in different quartiles. Additionally, there is no clear correlation between gdp quartile and mean latitude.